Development of a Reinforcement Learning Enabled Cattle Tracker Prototype

### 2025-02-13 paper/main.tex
Additional edits

There is an estimated 1.5 million beef cows in Alberta with a portion of those being free-range cattle who are allowed to graze on large pastures for extended periods of time.
Many farmers and researchers are interested in tracking movement of these cattle to asess the health of the herd, and to monitor the location of the cattle.
These trackers take the form of collars, or ear tags and are equipped with sensors, batteries, and communication modules.
The installation of the trackers is often a time consuming process so a tracker is expected to operate for extended periods of time (months) without human intervention.
Given the size of the devices and the environment that they are expected to operate in, there is a limited amount of power budget available to the device and it beocmes a balancing act to ensure contiunuous operation, transmission and data collection.
Reasearchers would like a device that transmits information at high rates as higher fidelity information allows them to make better conclusions about the state of the herd.
Sensors data that researchers are interested in includes GPS coordinates, accelerometer data snapshots, and temperature data.

The paradigm of Internet of Things (IoT) devices has emerged over the last two decade as a way to deploy compute and intelligence throughout the environment.
% The number of IoT devices is expected to grow to 75 billion by 2025 \cite{gubbiInternetThingsIoT2013}.
The devices are often small, low power, and have limited communication capabilities, but provide insights into the environment that they are deployed in.
The cattle trackers are an example of an IoT device that is deployed in the environment to provide insights into the health and location of the herd.
IoT devices combine sensors, communication modules, storage, and compute to provide insights into the environment that they are deployed in.
IoT devices come as networks or individual devices and in the case of a cattle tracker, ideally a tracker would be self-sufficient as a stand-alone item.
With the spread of the internet and Starlink-like services, it is possible to have a tracker that is connected to the internet and can send data to the cloud, which enables the farmer to monitor the herd from anywhere in the world.
Again we have the challenge of limited power budget and the need to transmit data at high rates.
The power budget can be extended with the use of solar power. 
Solar power is susceptible to weather conditions and the amount of sunlight that is available to the device, which in the case of cattle trackers also depends on the location/orientation of the cattle.

Reinforcement learning is a branch of machine learning that is allows an agent to learn optimum actions to take in an environment to maximize a reward.
The agent learns these actions by interacting with the environment and receiving feedback on the actions that it takes.
Reinforcement learning has been used in a variety of applications such as playing games, optimizing energy consumption, and optimizing the operation of IoT devices.
Machine learning deployed on IoT devices is somtimes referred to as Artificial intelligence of Things (AIoT)\cite{yamsaniIoTBasedLivestockMonitoring2024}, which includes a broad range of applications.
The TinyML paradigm is a subset of AIoT and focuses on machine learning on devices with highly constrained resources, which often limits it to inference only\cite{rayReviewTinyMLStateoftheart2022}.
On one hand the addition of intelligence to the device allows it to make better decisions, but on the other hand the addition of intelligence to the device increases the power consumption of the device.
Q-learning is a reinforcement learning algorithm that is used to optimize the actions that an agent takes in an environment and has been used in a variety of applications such as optimizing the operation of IoT devices.

In this paper we present the development of a self-sufficient cattle tracker prototype capable of monitoring a single cow for extended periods of time.
The tracker should record and collect data at regular intervals and transmit the data to the cloud with a maximum specified delay.
Solar power can be utilized to extend the power budget of the device.
Given that the device is capable of generating power, it should use excess power to transmit data more frequently, but the exact strategy is not trivial to determine.
Reinforcement learning is used to optimize the power consumption of the device by making decisions on when to transmit data, when to collect and store data, and when to do nothing.

@@

\section{Related Work}
% IoT and Cloud
The main challenges of IoT sensors are \cite{chenDeepReinforcementLearning2021} are (i) limited battery life, (ii) limited processing power, and (iii) limited storage.
IoT devices are often limited to inference only, as training a model on the device is computationally expensive.
Luckily, the cloud can be used to train the model and the model can be deployed on the device.

%Cattle
Cattle tracking with IoT devices is a growing field of research. In \cite{yamsaniIoTBasedLivestockMonitoring2024} the authors propose a network of IoT devices to monitor cattle.
The network consists of edge devices that are attached to the cattle, fog devices that are attached to the barn, and cloud devices that are attached to the cloud. 
In \cite{duttaMOOnitorIoTBased2022} the authors present an IoT device that clasifies the behaviour of the cattle based on data from accelerometer, temperature via XGBoost and Random Forest classifiers.
Their device lasts for an estimated 10 hours on a single charge.

% Reinforcement Learning
In \cite{hribarUsingDeepQLearning2019} the authors use deep Q-learning to extend the life of generic sensors networks with non-rechargable batteries. 
The authors take advantage of corelation between sensor measurements to determine the rate at which sensors should transmit to maximize battery life.

@@

>>>>

There is an estimated 1.5 million beef cows in Alberta with a portion of those being free-range cattle who are allowed to graze on large pastures for extended periods of time.
Many farmers and researchers are interested in tracking movement of these cattle to asess the health of the herd, and to monitor the location of the cattle.
Reasearchers are interested in behaviour of the herd like feeding, rumination, and resting and past research has shown that this behaviour can be correlated to sensor data from accelerometer \cite{unoldIoTBasedCowHealth2020}.
Reasearchers would like a device that transmits information at high rates as higher fidelity information allows them to make better conclusions about the state of the herd.
These trackers take the form of collars, leg bands, or ear tags and are equipped with sensors, batteries, and communication modules.
The installation of the trackers is a time consuming process so a tracker is expected to operate for extended periods of time (months) without human intervention.
Given the size of the devices and the environment that they operate in, there is a limited power budget available to the device and it beocmes a balancing act to ensure contiunuous operation, transmission and data collection.

The paradigm of Internet of Things (IoT) devices has emerged over the last two decades as a way to deploy sensors, communication modules, storage, compute, and intelligence throughout the environment.
% The number of IoT devices is expected to grow to 75 billion by 2025 \cite{gubbiInternetThingsIoT2013}.
The devices are often small, low power, and have limited communication capabilities, but provide insights into the environment that they are deployed in.
Cattle trackers are an example of an IoT device that is deployed in the environment to provide insights into the health and location of the herd.
These cattle trackers are strating to replace labour intensive manual observation with automated systems that can monitor the herd 24-7 \cite{unoldIoTBasedCowHealth2020} \cite{moutaouakilDigitalFarmingSurvey2023}.
IoT devices can operate as part of a network or as standalone units. 
For cattle tracking, a self-sufficient device that connects to the internet via cellular networks such as GSM, LTE, or even satellite services like Starlink is ideal, allowing farmers to monitor their herds remotely. 
However, these devices face power constraints due to the high data transmission rates required; solar panels can extend battery life, though their effectiveness depends on weather conditions and the cattle's orientation and location.

Reinforcement learning is a branch of machine learning that allows an agent to learn optimum actions to take in an environment to maximize a reward.
The agent learns these actions by interacting with the environment and receiving rewards for those actions.
Reinforcement learning has been used in a variety of applications such as playing games, optimizing energy consumption, and optimizing the operation of IoT devices.
The combination of reinforcement learning and IoT devices is an attractive option as it allows the device to make decisions without the need for human intervention.
THe combvination of AI and IoT is sometimes called the Artificial intelligence of Things (AIoT)\cite{yamsaniIoTBasedLivestockMonitoring2024}, which includes a broad range of applications.
The TinyML paradigm is a subset of AIoT and focuses on machine learning on devices with highly constrained resources, which often limits it to inference only\cite{rayReviewTinyMLStateoftheart2022}.
On one hand the addition of intelligence to the device allows it to make better decisions, but on the other hand the addition of intelligence to the device increases the power consumption of the device.
Q-learning is a reinforcement learning algorithm that is used to optimize the actions that an agent takes in an environment and has been used in a variety of applications such as optimizing the operation of IoT devices.

In this paper we present the development of a cattle tracker prototype capable of data collection for extended periods of time.
The main controbution of this paper is the development of a reinforcement learning agent that optimizes the power consumption of the device by allowing the agent to make decisions on when to transmit data, when to collect and store data, and when to do nothing.
The hardware prototype uses solar power to extend the power budget of the device and the agent is trained to take advantage of excess power availability to increase the data transmission rate.

@@

\section{Related Work}
% IoT and Cloud
The main challenges of IoT sensors are \cite{chenDeepReinforcementLearning2021} are (i) limited battery life, (ii) limited processing power, and (iii) limited storage.
Devices that fall under the TinyML paradigm are often limited to inference only, as training a model on the device is computationally expensive.
The addition of cloud communication opens up the possibility of training the model in the cloud and deploying a new model to the device that adapts to changes in the environment.

%Cattle
Agriculture and cattle tracking commonly employ IoT devices for monitoring. 
In \cite{yamsaniIoTBasedLivestockMonitoring2024} the authors propose a network of IoT devices to monitor cattle, with edge devices for network communication and cloud services for data storage and analysis.
Multiple standalone devices have also been developed for cattle tracking, with many focusing on the sensor output of the device.
In \cite{duttaMOOnitorIoTBased2022} the authors present an IoT device that clasifies the behaviour of the cattle based on data from accelerometer, temperature via XGBoost and Random Forest classifiers.
Their device lasts for an estimated 10 hours on a single charge.
In \cite{unoldIoTBasedCowHealth2020} the authors present the CowMonitor an IoT device uses Bluetooth Low Energy (BLE) for communication with a hub and sends accelerometer and magnetometer data to the cloud.
They transmit data without acknowledgement for oppurtunistic capture and lasts 5 years without battery replacement, but required the deployment of BLE hubs and WiFi ground stations thoughout the pasture.

% Reinforcement Learning
The combination of intelligence and IoT devices allows a powerful solution to add intelligence to the edge.
In \cite{hribarUsingDeepQLearning2019} the authors use deep Q-learning to extend the life of generic sensors networks with non-rechargable batteries. 
The authors take advantage of corelation between sensor measurements to determine the rate at which sensors should transmit to maximize battery life.

@@

It is not trivial to determine the reward values for the agent to mximize data collection.
Manual heuristics can be used to determine the reward values, but that is not efficient and may not be optimal.
For the agent we consider the reward values to be hyperparameters that can be tuned to maximize the data collection.
We use a Bayesian optimization framework—specifically, the Gaussian Process-based minimization function \verb|gp\_minimize| from the \verb|scikit-optimize| library to automatically search the hyperparameter space defined for each reward component (e.g., reward\_power\_loss, reward\_power\_multiplier, reward\_action\_0, reward\_action\_1, reward\_action\_2, and reward\_message\_count).
By repeatedly running training sessions with slightly varied random seeds and averaging the performance metrics, our objective function (which returns the negative median sent messages) guides the search toward the parameter set that maximizes data collection
This automated approach saves time compared to manual tuning efforts and has shown to be effective in optimizing the reward values.
